---
title: "abundance"
format: html
editor: visual
---

## Load Packages

```{r}
#| output: false
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(tidyverse, RPostgres, DBI, here, lubridate, unmarked)

```

## Connect to `survey_data` schema in `ribbitr` database

```{r}
#| output: false
tryCatch({
    drv <- dbDriver("Postgres")
    print("Connecting to Database…")
    connection <- dbConnect(drv,
                 dbname = Sys.getenv("aws_dbname"),
                 host = Sys.getenv("aws_host"),
                 port = Sys.getenv("aws_port"),
                 user = Sys.getenv("aws_user"),
                 password = Sys.getenv("aws_password"),
                 timezone=NULL)
    print("Database Connected!")
    },
    error=function(cond) {
            print("Unable to connect to Database.")
    })

#search path
dbExecute(connection, "set search_path to survey_data")

```

## Query SERDP Bd data - abundance

How do we want to handle day and night surveys?

```{r}

q_c <- "select r.region, s.site, v.date, v.survey_time,
     s2.detection_type, c.species_capture
     from region r
     join site s on r.region_id = s.region_id 
     join visit v on s.site_id = v.site_id 
     join survey s2 on v.visit_id = s2.visit_id
     join capture c on s2.survey_id = c.survey_id 
     where r.region = 'pennsylvania'
     and v.date < '2020-12-31';"

raw_cap <- dbGetQuery(connection, q_c)


q_v <- "select r.region, s.site, v.date
        from region r
        join site s on r.region_id = s.region_id 
        join visit v on s.site_id = v.site_id 
        where r.region = 'pennsylvania'
        and v.date < '2020-12-31'"

raw_visits <- dbGetQuery(connection, q_v) %>% 
  select(!region)

visits_per_site <- raw_visits %>% 
  mutate(year = year(date)) %>% 
  group_by(site) %>% 
  summarise(n = n())
```

## Summarize Data

```{r}

cleaned_up_cap <- raw_cap %>% 
  drop_na(species_capture) %>% 
  group_by(site, date, species_capture) %>% 
  summarise(n = n())


```

## Populate Site Zeros

```{r}

pop_zeros <- raw_visits %>% 
  left_join(cleaned_up_cap) %>% 
  complete(nesting(date, site),
           species_capture = unique(cleaned_up_cap$species_capture),
           fill = list(n = 0))

```

## practice

### yearly mean seasonal abundance estimates per site_year.

IE: tuttle_pond_2017, visit_1 = average count month 1 (march), visit_2 = average count month 2 (april), etc

```{r}

bull <- pop_zeros %>% 
  filter(species_capture == "lithobates_catesbeianus") %>% 
  mutate(year = year(date),
         site = paste(year, site, sep = "_"),
         month = month(date)) %>%
  group_by(month, site) %>% 
  mutate(n = sum(n)) %>% 
  ungroup() %>% 
  group_by(site) %>% 
  mutate(n_visit = match(month, unique(month)),
         n_visit = paste0("month_", month, sep = "")) %>% 
  ungroup() %>% 
  group_by(site, month, n_visit) %>% 
  mutate(temp_id = cur_group_id()) %>% 
  filter(!duplicated(temp_id)) %>% 
  ungroup() %>%
  select(!c(date, year, temp_id, month)) %>% 
  pivot_wider(names_from = c("n_visit"), values_from = c("n")) %>% 
  select(!c(site, species_capture)) 

bull_umf <- unmarkedFramePCount(y = bull)

fm1_bull <- pcount( ~1  ~1, 
               data = bull_umf, 
               engine = "R",
               control=list(trace=T, REPORT=5))

summary(fm1_bull)

backTransform(fm1_bull, "state")
backTransform(fm1_bull, "det")

```

### Mean Annual abundance estimates per site_year.

IE: tuttl, visit_1 = average count for 2017, visit_2 = average count for 2018, etc

This bad

```{r}

bull <- pop_zeros %>% 
  filter(species_capture == "lithobates_catesbeianus") %>% 
  mutate(year = year(date),
         #site = paste(year, site, sep = "_"),
         month = month(date)) %>%
  group_by(year, site) %>% 
  mutate(n = sum(n)) %>% 
  ungroup() %>% 
  group_by(site) %>% 
  mutate(n_visit = match(year, unique(year)),
         n_visit = paste0("year_", year, sep = "")) %>% 
  ungroup() %>% 
  group_by(site, year, n_visit) %>% 
  mutate(temp_id = cur_group_id()) %>% 
  filter(!duplicated(temp_id)) %>% 
  ungroup() %>%
  select(!c(date, year, temp_id, month)) %>% 
  pivot_wider(names_from = c("n_visit"), values_from = c("n")) %>% 
  select(!c(site, species_capture)) 

bull_umf <- unmarkedFramePCount(y = bull)

fm1_bull <- pcount( ~1  ~1, 
               data = bull_umf, 
               engine = "R",
               control=list(trace=T, REPORT=5))

summary(fm1_bull)

backTransform(fm1_bull, "state")
backTransform(fm1_bull, "det")

```