---
title: "abundance"
format: html
editor: source
---

any time point through time what is the abundance at each site or in Pennn fit gam 

## Load Packages

```{r}
#| output: false
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(tidyverse, RPostgres, DBI, here, lubridate, unmarked, mgcv, voxel, data.table)

```

## Connect to `survey_data` schema in `ribbitr` database

```{r}
#| output: false
tryCatch({
    drv <- dbDriver("Postgres")
    print("Connecting to Database…")
    connection <- dbConnect(drv,
                 dbname = Sys.getenv("aws_dbname"),
                 host = Sys.getenv("aws_host"),
                 port = Sys.getenv("aws_port"),
                 user = Sys.getenv("aws_user"),
                 password = Sys.getenv("aws_password"),
                 timezone=NULL)
    print("Database Connected!")
    },
    error=function(cond) {
            print("Unable to connect to Database.")
    })

#search path
dbExecute(connection, "set search_path to survey_data")

```

## Query SERDP Bd data - abundance

How do we want to handle day and night surveys?

```{r}

q_c <- "select r.region, s.site, v.date, v.survey_time,
     s2.detection_type, c.species_capture
     from region r
     join site s on r.region_id = s.region_id 
     join visit v on s.site_id = v.site_id 
     join survey s2 on v.visit_id = s2.visit_id
     join capture c on s2.survey_id = c.survey_id 
     where r.region = 'pennsylvania'
     and v.date < '2020-12-31';"

raw_cap <- dbGetQuery(connection, q_c)


q_v <- "select r.region, s.site, v.date
        from region r
        join site s on r.region_id = s.region_id 
        join visit v on s.site_id = v.site_id 
        where r.region = 'pennsylvania'
        and v.date < '2020-12-31'"

raw_visits <- dbGetQuery(connection, q_v) %>% 
  select(!region)

visits_per_site <- raw_visits %>% 
  mutate(year = year(date)) %>% 
  group_by(site) %>% 
  summarise(n = n())
```



## Summarize Data

```{r}

mid_clean_cap <- raw_cap %>% 
  drop_na(species_capture) %>% 
  group_by(site, date, species_capture) %>% 
  reframe(n = n())


```

## Populate Site Zeros
```{r}

clean_cap <- raw_visits %>% 
  left_join(mid_clean_cap, by = c("site", "date")) %>% 
  complete(nesting(date, site),
           species_capture = unique(mid_clean_cap$species_capture),
           fill = list(n = 0))

```

## Gam Model - all of Penn per species - assumming detection of 1 - We want raw counts on the plot or predicted points?
```{r}

# Create empty objects
species_list <- unique(clean_cap$species_capture)

plots <- list()

spec_names <- c()

# Loop through each species
for (i in 1:length(species_list)) {
  
  # find species with low k value
  k <- clean_cap %>% 
    filter(species_capture == species_list[i]) %>%
    mutate(year_month = as_date(date, format = "%Y-%m"),
           year_month = as.numeric(as.POSIXct(year_month, format = "%Y-%m")),
           year_month = sort(year_month)) %>%  
    select(year_month) %>% 
    unique() %>% 
    nrow()
  
  # only run models for species with k>8
  if (k > 9) {
  
      # add filter species list with k>8   
      #spec_names[i] <- species_list[i]
      
      # filter clean data set for species list of k>8
      dat <- clean_cap %>% 
        select(species_capture, date, n) %>% 
        filter(species_capture == species_list[i]) %>%  
        ungroup() %>% 
        select(!species_capture) %>% 
        mutate(unix = as_date(date, format = "%Y-%m"),
               unix = as.numeric(as.POSIXct(unix, format = "%Y-%m")),
               unix = sort(unix)) 

  
      # run GAM per species
      gam <- gam(n ~ s(unix, k = k), data = dat) 
      
      # create new object in environment for each model output per species
      assign(species_list[i], gam) 
      
      # Predictions
      unix = seq(min(dat$unix), max(dat$unix), len=100)
      newdata = data.frame(unix=unix)
      gam_pred = predict(gam, newdata=newdata, type="response", se.fit = T)
      newdata$pred = gam_pred$fit
      newdata$upper_se <- gam_pred$fit + (2 * gam_pred$se.fit)
      newdata$lower_se <- gam_pred$fit - (2 * gam_pred$se.fit)
      
      # save predictions
      write_rds(as.data.frame(newdata), here("outputs", "abundance", "gam_unix_per_species", paste(species_list[i],".rds", sep = "")))
      
      
      # Average prevalence for plotting
      dat = data.table(dat)
      dat[, c("month", "year"):=.(month(date), year(date))]
      average_n = dat[, .(av_n=mean(n), mean_unix=mean(unix)), by=.(month, year)] 

      # make GAM plot
      p <- ggplot(newdata) + geom_line(aes(x=unix, y=pred)) +
                             geom_point(data=dat, aes(x=unix, y=n), size=3, alpha = 0.5) + # raw or predicted counts displayed?
                             geom_ribbon(aes(ymin = lower_se, ymax= upper_se, x = unix), alpha = 0.2, fill = "blue")+
                             xlab("Unix") + ylab("Abundance (Perfect Detection)") + theme_classic() +
                             coord_cartesian(ylim = c(0, max(newdata$upper_se) + 5)) + ggtitle(species_list[i])
        
        
        
        
    # save GAM plot in list
     plots[[i]] <- p
    
  
  }
  
  # Error checking during loop
  print(paste("done with", species_list[i]))

}



```

## Examine plots

```{r}


plots[1:length(plots)]


```
## Save plots

```{r}
for (i in 1:length(plots)) {
  
  ggsave(here("figures", "abundance", "gam_unix_per_species", paste(species_list[i], ".jpeg", sep="")),
         plot = plots[[i]])
  
}



```

## Gam Model - per site per species - assumming detection of 1 - Nested for loop
```{r}
# Create empty objects
species_list <- unique(clean_cap$species_capture)

plots <- list()

all_plots <- list()

spec_names <- c()

site_list <- unique(clean_cap$site)

# Loop for site
for (j in 1:length(site_list)){

# Loop through each species
    for (i in 1:length(species_list)) {
      
      # find species with low k value
      k <- clean_cap %>% 
        filter(species_capture == species_list[i],
               site == site_list[j]) %>%
        mutate(year_month = as_date(date, format = "%Y-%m"),
               year_month = as.numeric(as.POSIXct(year_month, format = "%Y-%m")),
               year_month = sort(year_month)) %>%  
        select(year_month) %>% 
        unique() %>% 
        nrow()
      
      # only run models for species with k>8
      if (k > 9) {
      
          # add filter species list with k>8   
          spec_names[i] <- species_list[i]
          
          # filter clean data set for species list of k>8
          dat <- clean_cap %>% 
            select(species_capture, date, n, site) %>% 
            filter(species_capture == species_list[i],
                   site == site_list[j]) %>%  
            ungroup() %>% 
            select(!species_capture, site) %>% 
            mutate(unix = as_date(date, format = "%Y-%m"),
                   unix = as.numeric(as.POSIXct(unix, format = "%Y-%m")),
                   unix = sort(unix)) 
    
      
          # run GAM per species
          gam <- gam(n ~ s(unix, k = k), data = dat) 
          
          # create new object in environment for each model output per species
          assign(paste(site_list[j], "_",species_list[i], sep = ""), gam) 
          
          # Predictions
          unix = seq(min(dat$unix), max(dat$unix), len=100)
          newdata = data.frame(unix=unix)
          gam_pred = predict(gam, newdata=newdata, type="response", se.fit = T)
          newdata$pred = gam_pred$fit
          newdata$upper_se <- gam_pred$fit + (2 * gam_pred$se.fit)
          newdata$lower_se <- gam_pred$fit - (2 * gam_pred$se.fit)
          
          # save predictions
          #write_rds(as.data.frame(newdata), here("outputs", "abundance", "gam_unix_per_species", paste(species_list[i],".rds", sep = "")))
          
          
          # Average prevalence for plotting
          dat = data.table(dat)
          dat[, c("month", "year"):=.(month(date), year(date))]
          average_n = dat[, .(av_n=mean(n), mean_unix=mean(unix)), by=.(month, year)] 
    
          # make GAM plot
          p <- ggplot(newdata) + geom_line(aes(x=unix, y=pred)) +
                                 geom_point(data=dat, aes(x=unix, y=n), size=3, alpha = 0.5) + # raw or predicted counts displayed?
                                 geom_ribbon(aes(ymin = lower_se, ymax= upper_se, x = unix), alpha = 0.2, fill = "blue")+
                                 xlab("Unix") + ylab("Abundance (Perfect Detection)") + theme_classic() +
                                 coord_cartesian(ylim = c(0, max(newdata$upper_se) + 5)) + 
                                 ggtitle(paste(site_list[j], "-", species_list[i], sep = ""))
            
            
            
            
        # save GAM plot in list
         plots[[i]] <- p
        
      
      }
      
      # Error checking during loop
      #print(paste("done with", site_list[j], "_", species_list[i]))
    
    }
  
  all_plots[[i]] <- plots
}
```

## Examine plots

```{r}


plots[1:length(all_plots)]


```

## Examine Model Outputs

### Print all GAM model outputs - Make sure to run the above loop to populate correct GAM models

```{r}

ClassFilter <- function(x) inherits(get(x), 'gam')

Objs <- Filter(ClassFilter, ls())

print(Objs)

```
### Select which GAM model to view

```{r}
# Choose species from above list
summary(tryon_weber_plethodon_cinereus)
```



# Old N-Mix models

## Populate Site Zeros
```{r}

pop_zeros <- raw_visits %>% 
  left_join(cleaned_up_cap) %>% 
  complete(nesting(date, site),
           species_capture = unique(cleaned_up_cap$species_capture),
           fill = list(n = 0))

```

## practice

### yearly mean seasonal abundance estimates per site_year.

IE: tuttle_pond_2017, visit_1 = average count month 1 (march), visit_2 = average count month 2 (april), etc

```{r}

bull <- pop_zeros %>% 
  filter(species_capture == "lithobates_catesbeianus") %>% 
  mutate(year = year(date),
         site = paste(year, site, sep = "_"),
         month = month(date)) %>%
  group_by(month, site) %>% 
  mutate(n = sum(n)) %>% 
  ungroup() %>% 
  group_by(site) %>% 
  mutate(n_visit = match(month, unique(month)),
         n_visit = paste0("month_", month, sep = "")) %>% 
  ungroup() %>% 
  group_by(site, month, n_visit) %>% 
  mutate(temp_id = cur_group_id()) %>% 
  filter(!duplicated(temp_id)) %>% 
  ungroup() %>%
  select(!c(date, year, temp_id, month)) %>% 
  pivot_wider(names_from = c("n_visit"), values_from = c("n")) %>% 
  select(!c(site, species_capture)) 

bull_umf <- unmarkedFramePCount(y = bull)

fm1_bull <- pcount( ~1  ~1, 
               data = bull_umf, 
               engine = "R",
               control=list(trace=T, REPORT=5))

summary(fm1_bull)

backTransform(fm1_bull, "state")
backTransform(fm1_bull, "det")

```

### Mean Annual abundance estimates per site_year.

IE: tuttl, visit_1 = average count for 2017, visit_2 = average count for 2018, etc

This bad

```{r}

bull <- pop_zeros %>% 
  filter(species_capture == "lithobates_catesbeianus") %>% 
  mutate(year = year(date),
         #site = paste(year, site, sep = "_"),
         month = month(date)) %>%
  group_by(year, site) %>% 
  mutate(n = sum(n)) %>% 
  ungroup() %>% 
  group_by(site) %>% 
  mutate(n_visit = match(year, unique(year)),
         n_visit = paste0("year_", year, sep = "")) %>% 
  ungroup() %>% 
  group_by(site, year, n_visit) %>% 
  mutate(temp_id = cur_group_id()) %>% 
  filter(!duplicated(temp_id)) %>% 
  ungroup() %>%
  select(!c(date, year, temp_id, month)) %>% 
  pivot_wider(names_from = c("n_visit"), values_from = c("n")) %>% 
  select(!c(site, species_capture)) 

bull_umf <- unmarkedFramePCount(y = bull)

fm1_bull <- pcount( ~1  ~1, 
               data = bull_umf, 
               engine = "R",
               control=list(trace=T, REPORT=5))

summary(fm1_bull)

backTransform(fm1_bull, "state")
backTransform(fm1_bull, "det")

```
